---
title: "Correlations and Partial Correlations in R"
subtitle: ""
author: "Pablo E. Guti\u00E9rrez-Fonseca"
institute: ""
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, "css/nhsr.css", "css/nhsr-fonts.css", "css/custom.css"]
    nature:
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      highlightSpans: true 
      countIncrementalSlides: true
      ratio: "16:9"
    includes:
      after_body: [css/insert-logo.html]
xaringanExtra:
    use_panelset: true
---
```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readr)
library(lessR)
library(ggplot2)
library(palmerpenguins)
library(car)
library(RVAideMemoire) #shapiro.test
xaringanExtra::use_panelset()
```

---
```{r figure_1, echo = FALSE, fig.align='center'}
DiagrammeR::mermaid("
    graph TD;

        A[What type of data?] --> B[Continuous] 
        A --> C[Categorical]
        
        B --Research Question--> D[Comparing Differences] 
        B --Research Question--> E[Examining Relationships]
        
        D --> F[How many groups?] 
      
        F --1 group--> G[Normally distributed?] 
        G --Yes--> H[One-sample z-test] 
        G --No--> I[Wilcoxon signed rank test]
        
        F --2 groups--> J[Are Samples Independent?] 
        J --Yes--> K[Normally distributed?]
        J --No--> L[Normally distributed?]
        
        K --Yes--> M[Independent t-test] 
        K --No--> N[Wilcoxon rank sum test] 
        
        L --Yes--> O[Paired t-test] 
        L --No--> P[Wilcoxon matched pair test]
        
        F --More than 2 groups--> Q[Number of Treatments?]
        
        Q --One--> R[Normally Distributed?]
        R --Yes--> S[One-Way ANOVA]
        R --No--> T[Non-Parametric ANOVA]
        
        Q --More than One--> U[Normally Distributed?]
        
        U --Yes--> V[Factorial ANOVA]
        U --No--> W[Friedman's Rank Test]
        
        E --> X[Number of Treatment Variables?]
        
        X --Two--> Y[Normally Distributed?]
        Y --Yes--> Z[Pearson's Correlation]
        Y --No--> AA[Spearman's Rho]
        
        X --More than Two--> AB[Partial Correlation]

  style A fill:lightblue,stroke:#333,stroke-width:1px
  style B fill:lightblue,stroke:#333,stroke-width:1px
  style C fill:lightblue,stroke:#333,stroke-width:1px
  style D fill:lightblue,stroke:#333,stroke-width:1px
  style E fill:lightblue,stroke:#333,stroke-width:1px
  style F fill:lightblue,stroke:#333,stroke-width:1px
  style G fill:lightblue,stroke:#333,stroke-width:1px
  style H fill:lightblue,stroke:#333,stroke-width:1px
  style I fill:lightblue,stroke:#333,stroke-width:1px
  style J fill:lightblue,stroke:#333,stroke-width:1px
  style K fill:lightblue,stroke:#333,stroke-width:1px
  style L fill:lightblue,stroke:#333,stroke-width:1px
  style M fill:lightblue,stroke:#333,stroke-width:1px
  style N fill:lightblue,stroke:#333,stroke-width:1px
  style O fill:lightblue,stroke:#333,stroke-width:1px
  style P fill:lightblue,stroke:#333,stroke-width:1px
  style Q fill:lightblue,stroke:#333,stroke-width:1px
  style R fill:lightblue,stroke:#333,stroke-width:1px
  style S fill:lightblue,stroke:#333,stroke-width:1px
  style T fill:lightblue,stroke:#333,stroke-width:1px
  style U fill:lightblue,stroke:#333,stroke-width:1px
  style V fill:lightblue,stroke:#333,stroke-width:1px
  style W fill:lightblue,stroke:#333,stroke-width:1px
  style X fill:lightblue,stroke:#333,stroke-width:1px
  style Y fill:lightblue,stroke:#333,stroke-width:1px
  style Z fill:lightblue,stroke:#333,stroke-width:1px
  style AA fill:lightblue,stroke:#333,stroke-width:1px
  style AB fill:lightblue,stroke:#333,stroke-width:1px

  ", height = '600px', width = '900px'
)


```

---
# What is Correlation?
- A correlation test quantifies how two variables change together.
  - Are the variables closely related?
  - To what extent does one variable predict the other?

---
# When to Use Correlation?

- You have two continuous variables measured from the same observational unit.
- You want to determine if the variables are related (vary in unison).

---
# Key Assumptions:
- Independent, random samples.

- Normality:
  - Pearson correlation (parametric): Assumes both variables follow a normal distribution.
  - Spearman correlation (nonparametric): Does not assume normality.

- Linear relationship: Correlation looks for a linear association between the variables.

---
# Describing Correlations
- Strength: Indicates how closely the variables are related.
    - Categories: Weak, Moderate, Strong
    
- Nature: Shows the direction of the relationship.
    - Direct (Positive): As one variable increases, so does the other.
    - Indirect (Negative): As one variable increases, the other decreases.

- Significance: Assesses if the correlation is meaningful.
    - Based on the p-value, which tells us if the observed relationship is likely due to chance.
