---
title: "Diversity Indices"
author: "Pablo E. Gutiérrez-Fonseca"
date: "2023-10-01"
output: pdf_document
---


## **Load libraries**
```{r}
library(vegan)
```


```{r}
data(BCI, BCI.env)
```

## **Species richness**<br/>
Species richness is a measure of the number of species (or other taxonomic level) present at a site.<br/>
Sites with more taxa are considered **richer**.

## **Diversity indices**

### **Shannon_Weaver (or H)**

The Shannon entropy (\(H\)) is calculated using the formula:  

\[ H = -\sum_{i=1}^{S} p_i * \ln(p_i) \]  

where:  
- \(\sum\): A Greek symbol that means "sum",  
- \(\ln\): Natural logarithm,  
- \( p_i \): The proportion of the entire community made up of species \(i\).  

This formula quantifies the uncertainty or information content in a system with multiple species, 
where each \( p_i \) represents the relative abundance of a specific species.
- The higher the value of H, the higher the diversity of species in a particular community.  
- The lower the value of H, the lower the diversity.  
- A value of H = 0 indicates a community that only has one species.


