---
title: "Diversity Indices"
author: "Pablo E. Gutiérrez-Fonseca"
date: "2023-10-01"
output: pdf_document
---


## **Load libraries**
```{r}
library(vegan)
library(dplyr)
library(readxl)
```
<br/>
<br/>

## **Aquatic Macronvertebrates in two land uses (Agriculture and Urban streams)**

```{r}
data <- read_excel('data.xlsx', sheet = 'Sheet3')
head(data,20)
```
<br/>
<br/>

## **Species richness**<br/>
Species richness is a measure of the number of species (or other taxonomic level) present at a site.<br/>
Sites with more taxa are considered **richer**.
```{r}
# Calculate total species richness per site
total_species_richness <- data %>%
  group_by(site) %>%
  summarise(total_species = n_distinct(taxa))

# Display the results
print(total_species_richness)
```

## **Abundance**
```{r}
# Calculate mean abundance per taxa across all sites
mean_abundance <- data %>%
  group_by(site) %>%
  summarize(mean_abundance = sum(abundance, na.rm = TRUE))

# Display the result
print(mean_abundance)
```

## **Diversity indices**

### **Shannon_Weaver (or H)**

The Shannon entropy (\(H\)) is calculated using the formula:

\[ H = -\sum_{i=1}^{S} p_i * \log_b(p_i) \]

where:  
- \( p_i \) is the proportion of species \(i\),  
- \( S \) is the total number of species,  
- \( \sum_{i=1}^{S} p_i = 1 \) (the sum of all proportions is equal to 1),  
- \( b \) is the base of the logarithm.

This formula is commonly used to measure the diversity or uncertainty in a system with multiple species, where \( p_i \) represents the relative abundance of each species.  
- The higher the value of H, the higher the diversity of species in a particular community.  
- The lower the value of H, the lower the diversity.  
- A value of H = 0 indicates a community that only has one species.


## Example
```{r}
# Calculate Shannon diversity index
shannon_results <- data %>%
  group_by(site) %>%
  summarise(shannon_index = diversity(c(abundance), index = "shannon"))

# Display the results
print(shannon_results)
```




