---
title: "Introduction to R"
author: 
- "Pablo E. Gutiérrez-Fonseca"
- "`r format(Sys.time(), '%d %B, %Y')`"
layout: default
permalink: /pages/Lectures/Introduction to R/
output: html_document
---
<br/>
<br/>

#### What is R?
Is a computer language that allows the user to program algorithms and use tools 
that have been programmed by others. 
R is based on the computer language S, which was developed by John
Chambers and others at Bell Laboratories in 1976. In the early 1990s, Ross
Ihaka and Robert Gentleman (University of Auckland, in Auckland, New
Zealand) experimented with the language and called their product R. Note
that both their first names begin with the letter R. Also when modifying a
language called S, it is tempting to call it T or R.

R comes with a series of default packages. A package is a collection of previously programmed functions, often including functions for specific tasks. It is
tempting to call this a library, but the R community refers to it as a package.

<br/>

#### Change appearance
Tools -> Global options... -> Appearance 

<br/>

#### >
Significa "What now?"
<br/>The screen prompt ">" is an invitation to put R to work.
<br/>Each line can have at most 128 characters.
<br/>When continuation is expected, the prompt changes from > to +
<br/>Note that the + continuation prompt does not carry out arithmetic plus.

<br/>

#### How cite R?
```{r}
citation()
```

<br/>

#### Cite specific packages?
```{r}
citation('vegan')
citation('ggplot2')
citation('randomForest')
```

<br/>

#### Help
```{r, eval=FALSE}
?lm
```

<br/>

#### Simbols (P: 22)
```{r, eval=FALSE}
+ - * / ^ arithmetic
> >= < <= == != relational
! & logical
~ model formulae
<- -> assignment
$ list indexing (the ‘element name’ operator)
: create a sequence
```

<br/>

#### Equality
```{r}
TRUE==TRUE
'HELLO'=='GOOBYE'
'HELLO'=='HELLO'
3==2
3==3
```

<br/>

#### Inequality
```{r}
TRUE!=TRUE
'HELLO'!='GOOBYE'
```

<br/>

#### Change directory
```{r, eval=FALSE}
setwd ("C:/users/Pablo Guti/desktop/")
```

<br/>

#### If you want to find out the name of the current working directory, use:
```{r}
getwd()
```

<br/>

#### Instals packages
```{r, eval=FALSE}
install.packages()
install.packages()
install.packages()
```

<br/>

#### Instals libraries
```{r, eval=FALSE}
library( )
library( )
library( )
```

<br/>

#### Exercises 
```{r}
5+6+3+6+4+2+4+8+3+2+7
2+3; 5*7; 3-7
```


```{r}
pi
sin(pi/2)
cos(pi/2)
```

<br/>

### Exponentes
```{r}
1.2e3
1e4
3e4
```

<br/>

#### Redondear 
```{r}
floor(5.7)
ceiling(5.7)
```

<br/>

####  c()function, where c stands for concatenate
```{r}
Wingcrd <- c(59, 55, 53.5, 55, 52.5, 57.5, 53, 55)
Wingcrd
Wingcrd [3]
Wingcrd [1 : 5]
Wingcrd [-2]

sum(Wingcrd)
s.win <- sum(Wingcrd)
s.win
```

<br/>

```{r}
Tarsus <- c(22.3, 19.7, 20.8, 20.3, 20.8, 21.5, 20.6,21.5)
Head <- c(31.2, 30.4, 30.6, 30.3, 30.3, 30.8, 32.5,NA)
Wt <- c(9.5, 13.8, 14.8, 15.2, 15.5, 15.6, 15.6,15.7)

Bird<-c(Tarsus,Head,Wt)
Bird
```

<br/>

#### cbind()function, combines the variables in such a way that the output
#### contains the original variables in columns. 
```{r}
Bird1 <- cbind(Tarsus, Head, Wt)
Bird1
```

<br/>

###  similar to the cbind function to arrange the variables in columns, 
###  the rbind function combines the data in rows. 
```{r}
Bird2 <- rbind(Tarsus, Head, Wt)
Bird2
```

<br/>

#### Remover NA
```{r}
x<-c(1:8,NA)
mean(x)
x
```

<br/>

```{r}
Head <-c(31.2,30.4,30.6,30.3,30.8,NA,32.5,NA)
Head
sum(Head)
sum(Head, na.rm=TRUE)
```

<br/>

####   na.rm=TRUE 
```{r}
mean(x,na.rm=T)
```

<br/>

####   Convertir NA en Ceros
```{r}
vmv<-c(1:6,NA,NA,9:12)
vmv

which(is.na(vmv))

vmv[is.na(vmv)]<- 0
vmv
```

<br/>

#### Asigna un determinado elemento
```{r}
A<-1:10
A

B<-c(2,4,8)
B

mean(A)
mean(B)
max(B)
min(A)
sum(A)
```

<br/>

#### If you want to begin a new analysis on a different dataset, it may be useful to
#### remove all variables. One option is to quit R and restart it. Alternatively,
```{r}
rm(list = ls(all = TRUE))
```

<br/>

#### Import data (.csv, .txt, direct import)
```{r, eval=FALSE}
xxx.frm=read.table("xxx.txt",header=T)
attach(xxx.frm)
xxx.frm

xxx.frm=read.csv("xxx.csv")
attach(xxx.frm)
xxx.frm

data<-read.table(file.choose(),header=T)
```

<br/>

#### References:

Crawley, M. J. (2012). The R book. John Wiley & Sons.

<br/>
<br/>
<br/>
