---
title: "M3 Problem Set"
output: pdf_document
date: "2024-02-05"
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```


### Loading libraries
library(tidyverse)  
library(palmerpenguins)  
library(ggthemes)  
```{r, include=FALSE}
library(tidyverse)
library(palmerpenguins)
library(ggthemes)
```


### Loading
```{r}
data(trees)
data(palmerpenguins)
```

### Print the first 6 rows
```{r}
head(trees)
head(penguins)
```


## Questions
1. Using the **_nrow()_** command, how many rows does the data.frame **_tress_** have?
```{r}
nrow(trees)
```


2. Using the **_ncol()_** command, how many columns are there in the **trees** dataset?
```{r}
ncol(trees)
```


3. What information does the **_length(trees)_** command provide about the structure of the trees data frame in R?
```{r}
length(trees)
```


4. Create a vector with elements **_1 2 3 4 5 6 7 8 9_** and call it x.
create another vector with elements **_10 20 30 40 50_** and call it y.
What happens if you try to add **_x_** and **_y_** together? why?
```{r}
x <- c(1, 2, 3, 4, 5, 6, 7, 8, 9)
y <- c(10, 20, 30, 40, 50, 60, 70, 80)
x+y
```


5. Data Visualization **_[link to section here](https://r4ds.hadley.nz/data-visualize)_**.
Go through section 1.2.  
This section involves walking through building a ggplot from the ground up, using a dataset which is built into R already.   
There are a number of datasets which come with R, which you can access
through the datasets library. (Try running library(help = "datasets")in R to see a
list). Make sure to do this section step by step to avoid encountering errors! (you still
may encounter errors, which again, is normal and is a prompt for you to understand
what is happening).  

Your submission for this question should be the graph of the penguin data. 
a. Which of our ﬁsh species seems to have a higher length for a given weight compared to
the other ﬁsh? How do you know? Do a google image search for the species to see if your
answer makes sense!
