---
title: "range shift"
output: pdf_document
date: "2025-03-05"
---

Chen et al. (2011) wanted to test the idea that organisms move to higher 
elevation as the climate warms. To test this, they collected data from 31 species, 
plotted below.

```{r, message= FALSE, warning=FALSE}
library(readxl)
library(BSDA)
library(dplyr)
library(ggplot2)
```


```{r}
df <- read.csv('range_shift_data.csv')
head(df)
```

# Plot
```{r}
range_shift <- df |> 
  mutate(x = "", uphill = elevationalRangeShift > 0)

ggplot(range_shift, aes(x = x, y = elevationalRangeShift))+
  geom_jitter(aes(color = uphill), width = .05, height = 0, size = 2, alpha = .7)+
  geom_hline(yintercept = 0, lty= 2)+
  stat_summary(fun.data = "mean_cl_normal") + 
  theme(axis.title.x = element_blank())

```


# Normality test
```{r}
shapiro.test(df$elevationalRangeShift)
```


# Estimate population parameters
```{r}
mean_pop <- 0
sd_pop <- sd(df$elevationalRangeShift)
```


# Run z.test
```{r}
z.test(df$elevationalRangeShift,
       alternative='greater',
       mu=mean_pop,
       sigma.x=sd_pop)
```


# Reference
Chen, I-Ching, Jane K. Hill, Ralf Ohlemüller, David B. Roy, and Chris D. Thomas. 
2011. Rapid Range Shifts of Species Associated with High Levels of Climate Warming. Science 333 (6045): 1024 26. https://doi.org/10.1126/science.1206432.
