---
title: "Experimental Design"
subtitle: ""
author: "Pablo E. Guti\u00E9rrez-Fonseca"
institute: ""
date: "2025-01-15 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, "css/nhsr.css", "css/nhsr-fonts.css", "css/custom.css"]
    nature:
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      highlightSpans: true 
      countIncrementalSlides: true
      ratio: "16:9"
    includes:
      after_body: [css/insert-logo.html] #this is for logos in all slides
xaringanExtra:
    use_panelset: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(readr)
library(lessR)
library(ggplot2)
library(patchwork)
library(palmerpenguins)
library(car)
library(ggforce) # for geom_circle
library(RVAideMemoire) #shapiro.test
library(DiagrammeR)
knitr::opts_chunk$set(dpi= 300)
xaringanExtra::use_panelset()
xaringanExtra::use_scribble()
xaringanExtra::use_search(show_icon = FALSE, position= "bottom-left") # Search
xaringanExtra::use_progress_bar(color = "#0051BA", location = "bottom", 
                                height = "4px")
xaringanExtra::use_clipboard() # Copy Code 
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)
xaringanExtra::use_editable(expires = 1) # Add textboxes to edit during presentation
```

# What is Experimental Design?

- An experimental design is the laying out of a detailed experimental plan in advance of doing the experiment. 

- Why it is important?
    - Well chosen experimental design maximize the amount of information that can be obtained for a given amount of experimental effort. 
    - Without proper design, inferential statistics is something not even possible. 

---
# Why should you care?
- Without proper design it is often impossible to get the information you want from data collected.

- It will save you and time and money.

- Employers know this.. And know they need people who can do it.

---
# Key Information for Experimental Design
1. Define the research question: Hypothesis or study objectives.

2. State variables of interest: Independent and dependent variables, controls, covariates.

3. Construct a sampling design:
    - Define the population of interest.
    - Determine the sampling unit: What is an observation?
    - Appropriate sampling unit size and shape.
    - How many observations are needed?
    - Selected and position of observations: consider control and pairing.

4. Develop your **method**: how, where, when to sample?

- This information will help you to choose the appropriate statistical analysis.
     - We'll cover this throughout the course.

---
# Define the problem: Hypothesis or study objective
- Hypothesis: statement of the expected outcome of an experiment.

- Characteristics of a good Hypothesis:
    - Clear, concise statement of expected results.
    - Based on scientific theory or observation
    - Is testable
    - Identifies the key variables included 
    - Specifies the nature of the relationship or differences expected
    - Define the population of interest.

---
# Hypothesis or study objective

.pull-left[ 
- Clear and Concise: Clearly states the expected outcome.
    - **Example:** Higher soil moisture will increase tree growth rates in Vermont forests.

- Based on Theory or Observation: Grounded in prior knowledge or observations.

.medium[
    - **Example:** Based on the understanding that moisture availability influences tree growth.]

- Testable: Can be evaluated with data collection.
    
.medium[
    - **Example:** Measure tree growth in areas with varying soil moisture levels.]
]

.pull-right[
- Identifies Key Variables: Specifies independent and dependent variables.

.medium[
    - **Example:** Soil moisture (independent) and tree growth rate (dependent).]

- Specifies Relationship: Defines the expected interaction between variables.

.medium[
    - **Example:** Increased soil moisture is expected to enhance tree growth.]

- Defines Population: Specifies the study context or target population.

.medium[
    - **Example:** Mature deciduous trees in Vermont forests.]
]

---
# Hypothesis or **study objective**

- A study objective outlines the purpose of research when there is no specific hypothesis to test statistically. Instead, it focuses on describing, monitoring, or understanding a system. Here are the key characteristics:

1. Clear and Concise Statement of the Goal or Knowledge to Be Gained:  
The objective should clearly articulate what the study aims to achieve or learn.
.medium[
    - **Example:** To assess the effects of varying soil moisture on tree growth in Vermont forests.]

2. Identified Key Variables:  
Clearly identify the main variables that will be measured or observed in the study.
.medium[
    - **Example:** Soil moisture levels and tree growth rates.]

3. Defines the Population of Interest:  
.medium[
    - Specify the specific population or system that the study will focus on.]

---
# Variables of interest

- Independent Variable (Manipulated Variable):
    - This is the variable that is deliberately changed or manipulated in the study. It is the primary focus of the experiment to see how it affects the dependent variable.
    - Example: In a study on tree growth, the independent variable could be soil moisture levels.

- Dependent Variable (Responding Variable):
    - This variable changes in response to the manipulation of the independent variable. It is what you measure to assess the effect of the independent variable.
    - Example: The growth rate of trees, which responds to changes in soil moisture.
    
---
# Variables of interest

- **Control:**
 - Controls are factors that are kept constant throughout the experiment to ensure that the results are due to the manipulation of the independent variable, not other factors. Controls help establish a standard of comparison.
 - Example: Using the same tree species, consistent light exposure, and the same soil type across all test groups.

- **Covariate:**  
- Covariates are factors that could influence the dependent variable but cannot be fully controlled. Including them helps to account for their potential impact in statistical analysis, ensuring more accurate results.
- Example: In a tree growth study, covariates might include temperature, nutrient availability, or pest pressure-factors that can influence growth but are difficult to control completely.

---
# Sample, replicate and pseudo replicate 
- **Sample:**
- A statistical sample is a subset of items which is used to represent the population from which they are drawn.

- It has two connotations (Sokal & Rohlf 2009):
1. **Individual observations**, which are taken in small units.
These smaller sampling units are often, but not necessarily, individuals in the biological sense.

If we take the weight of 100 fish, then the weight of each fish is an individual observation.

---
# Sample, replicate and pseudo replicate 
- **Sample:**
- A statistical sample is a subset of items which is used to represent the population from which they are drawn.

- It has two connotations (Sokal & Rohlf 2009):
2. **A set of individuals** that are examined to estimate the average value.
Each set of individuals are considered as basic sampling units.

- If we wish to measure temperature in a study of ant colonies, where each colony is a basic sampling unit, each temperature reading for one colony is an individual observation, and the sample of observations is the temperature for all the colonies considered.


---
# Sample, replicate and pseudo replicate 

- **Replication:** 
- Replication means having observations replicated at a spatial and temporal scale that coincides with the application of the experimental treatments.
- In statistics, replicates mean different samples that capture random variations of the system.
- Replicas are essential because biological systems are inherently variable, and this is particularly true for ecological systems.

---
# Sample, replicate and pseudo replicate

- **Pseudoreplication:**
Hurlbert (1984) to refer to **the use of inferential statistics to test for treatment effects with data from experiments where either treatments are not replicated (though samples may be) or replicates are not statistically independent.**

Heffner et al (1996) distinguish a pseudoreplicate from a true replicate, which they characterize as **the smallest experimental unit to which a treatment is independently applied**.

---
# Sampling design


---
# Sampling design

- Sampling design proposed method of selecting a subset of the population for measurements.

- Observation (or Sampling Unit):
  - Individual Plants or Animals: Measure specific organisms.
  - Plant or Animal Parts: Sample parts (e.g., leaves, scales).
  - Quadrants: Define and sample areas (e.g., 1m^2).
  - Lines (Transects): Sample along a line across the study area.
  - Points: Collect samples at specific points.
  
---
# Understanding Statistical Power
- **Statistical Power** is the probability that a statistical test will correctly reject a null hypothesis (which states that there is no relationship between the variables of interest) when it is false, meaning that a true effect or relationship exists. 
It reflects the ability of the test to detect a significant result if there is one.

- Power Value: A power of 0.9 means there is a 90% chance of detecting a significant difference if it truly exists, and a 10% chance of missing it (Type II error).
For most studies, a power greater than 0.8 is considered adequate, indicating an 80% chance of detecting a significant result if there is a true effect.

- Statistical power is intricately linked to the number of observations.

---
# Understanding Statistical Power

---
# Sampling design

.center[

```{r, echo=FALSE}
mermaid("
graph TB
    A[Sampling techniques]
    A --> B[Non-Probability sampling<br>Chosen based on subjective judgment]
    A --> C[Probability sampling, is a technique in which the members of the sample<br> are chosen randomly from the population.<br> Every member has an equal opportunity to be a part of a sample of this type. ]
",
height = '100%', width = '100%')
```

]

---
# Simple Random Sampling

.center[

```{r, echo=FALSE}
mermaid("
graph TB
    A[Sampling techniques]
    A --> B[Non-Probability sampling<br> 1. Judgmental]
    A --> C[Probability sampling<br> 1. Simple random sampling<br> 2. Systematic and Grid sampling<br> 3. Stratified sampling<br> 4. Adaptive Cluster samplingNbr> 5. Ranked Set Sampling]
",
height = '100%', width = '100%')
```

]

---
# Judgmental Sampling

- **Definition:** Selection of sampling units (number, location, timing) based on the researcher's judgment rather than random or systematic methods.

- **Limitations:** Conclusions are limited by the accuracy and validity of the researcher's judgment. Probabilistic statements about parameters are not possible.

- **Appropriate Use:**
  - Descriptive Studies: To explore or describe phenomena where random sampling is impractical.
  - Pilot Studies: To test methods or gather initial insights.
Emergency Response: For immediate data collection when quick decisions are needed.

---
# Judgmental Sampling
- You are monitoring the wolf population in Yellowstone and need to quickly estimate breeding success to determine if further monitoring is necessary. You have time to assess 3 of your packs.

- Random Sampling: Of the 10 packs in the park, you randomly select 3 to visit.
Judgmental Sampling (Vulnerable Packs): You visit the dens of the most vulnerable packs within the park's borders to assess litter size.

- Judgmental Sampling (Least Impacted Packs): You visit the dens of the packs that you know are least impacted by your intrusion to assess litter size.

---
# Judgmental Sampling

- You are monitoring water quality in Lake Champlain but have time to collect only 10 samples at regular intervals.

- Random Sampling: Use GPS to randomly select 10 points around the Lake Champlain waterfront.

- Judgmental Sampling: Collect samples from 10 key locations where pollution loads are highest (e.g., near sewage plants, stormwater drains, or areas with agricultural runoff).

---
# Simple Random Sampling

.center[

```{r, echo=FALSE}
mermaid("
graph TB
    A[Sampling techniques]
    A --> B[Non-Probability sampling<br> 1. Judgmental]
    A --> C[Probability sampling<br> 1. Simple random sampling<br> 2. Systematic and Grid sampling<br> 3. Stratified sampling<br> 4. Adaptive Cluster sampling<br> 5. Ranked Set Sampling]
",
height = '100%', width = '100%')
```

]

---
# Simple Random Sampling
- **Definition:** Applies sampling theory with random selection, giving each member of the population an equal chance of being selected.

- **Benefits:**
  - Provides statistically unbiased estimates of the mean and variability of the larger population.
  - Enables probabilistic statements about parameters, supporting conclusions with confidence levels.
  - Allows for statistical inference.

- **Applications:** Ideal for comparative, experimental, and modeling studies.

---
# Simple Random Sampling
- Key Assumption of Probability-Based Design: 
    - The sample must be representative of the entire population.

- To achieve this: 
    - Use random, unbiased sampling methods.
    - Collect enough samples to ensure accurate representation.
    - Distribute sampling units well across the population.
    - Ensure sampling units are independent (no spatial correlation).
    

---
# Simple Random Sampling
- Simple Random Sampling: Sampling units are selected randomly, with each having an equal chance of being chosen.

- Best Use: Ideal when the population is relatively homogeneous, with no major patterns or hotspots expected.


---
# Stratified Random Sampling
- **Stratified Random Sampling:** The target population is divided into non-overlapping strata or subpopulations that are more homogeneous.

- **Best Use:** 
Effective when the population is heterogeneous and can be subdivided by specific characteristics. It enhances precision in estimating means and variances and provides reliable estimates for subgroups of special interest.

---
# Systematic and Grid Sampling
- **Systematic and Grid Sampling:** An initial location or time is chosen at random, followed by sampling at regular intervals across space or time.

- **Best Use:** Useful for identifying hotspots, estimating spatial patterns, and detecting trends over time. Ensures uniform coverage of the study area.

---
# Adaptive Cluster Sampling
- **Adaptive Cluster Sampling:** Starts with simple random sampling, then additional samples are taken where measurements meet specific criteria, with potential for several rounds of sampling and analysis.

- **Best Use:** Ideal for estimating or detecting rare characteristics and is suited for inexpensive, rapid measurements.  It helps delineate hotspot boundaries while providing unbiased estimates of the population mean using appropriate weighting of all collected data.

---
# Ranked Set Sampling

- **Ranked Set Sampling:** Uses a two-phase design where inexpensive measurements rank multiple locations within a set, then selects one location from each set for more detailed sampling.

- **Best Use:** Cost-efficient and incorporates professional judgment into a random design. It typically provides more representative samples and more precise estimates of the population mean compared to simple random sampling.


---
# Key Information for Experimental Design

- Since 2004 the number of taps in Vermont has gone from a reported 2,100,000 to 4,490,000 and production has gone from 500,000 gallons to 1,390,000 gallons the value of this production has risen from approximately 14 million dollars in 2004 to 44.5 million in 2014 (National Agricultural Statistics Service, 2006 and 2015). 

---
# Key Information for Experimental Design

1. Define the problem:  Hypothesis or study objective: 
    - Does industrial scale sugaring impact the health of sugar maple stands.

--

2. Identify the population of interest (what are we trying to represent)
    - Sugar maple stands in Vermont.

--

3. State Variables of interest (independent, dependent, controls, covariates)
    - Dependent:  Mean Crown Vigor (0-4, healthy to dead)
  - Independent:  Sugaring intensity (none, low, high)
  - Confounding factors:  Soil type, elevation, drainage, stand density, disturbance (0/1)
  
---
# Key Information for Experimental Design
- Construct a Sampling Design
    - Determine the Sampling unit (what is an observation) 
    - Appropriate sampling unit size and shape?

- Let's follow best practices and use the FIA (Forest Inventory Analysis) inventory and assessment plot design.
    - Each plot consists of 4 sub-plots.

    - Sampling Unit = plot
    - Sub-plots = replicates
    
---
# Sampling Design

- You can assess canopy condition at up to 30 plots across the state.

- Each of these plots is made up of 4 sub-plots
    - A plot is a sampling unit
    - A subplot is a replicate within each plot
    - But how do I pick where to put these 30 plots?

---
# Sampling Design

.pull-left[ 
- **Simple Random Sampling:** 
    - Randomly select 30 plots across the state.
    - Each plot consists of 4 sub-plots.
    - Best for homogeneous populations with no expected patterns.
    
]
.pull-right[
]

---
# Sampling Design
- **Stratified Random Sampling:** 
    - Divide the state into regions based on Sugaring operations.
    - Randomly select 10 plots from each stratum.
    - Each plot consists of 4 sub-plots.
    - Best for heterogeneous populations with distinct subgroups.

---
# Sampling Design
- **Systematic and Grid Sampling:** 
    - Create a grid across the state.
    - Randomly select 30 grid points.
    - Each plot consists of 4 sub-plots.
    - Best for identifying hotspots or trends across space.
    
---
# Sampling Design
- **Adaptive Cluster Sampling:** 
    - Randomly select 30 plots across the state.
    - Assess canopy condition.
    - Select additional plots based on canopy condition.
    - Each plot consists of 4 sub-plots.
    - Best for identifying rare characteristics or hotspots.
    
---
# Sampling Design
- **Judgmental Sampling:** 
    - Select 30 plots based on expert knowledge.
    - Each plot consists of 4 sub-plots.
    - Best for exploratory studies or when quick decisions are needed.
    
    


