---
title: "Experimental Design"
subtitle: ""
author: "Pablo E. Guti\u00E9rrez-Fonseca"
institute: ""
date: "`r Sys.Date()`"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, "css/nhsr.css", "css/nhsr-fonts.css", "css/custom.css"]
    nature:
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      highlightSpans: true 
      countIncrementalSlides: true
      ratio: "16:9"
    includes:
      after_body: [css/insert-logo.html] #this is for logos in all slides
xaringanExtra:
    use_panelset: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(readr)
library(lessR)
library(ggplot2)
library(patchwork)
library(palmerpenguins)
library(car)
library(ggforce) # for geom_circle
library(RVAideMemoire) #shapiro.test
library(DiagrammeR)
knitr::opts_chunk$set(dpi= 300)
xaringanExtra::use_panelset()
xaringanExtra::use_scribble()
xaringanExtra::use_search(show_icon = FALSE, position= "bottom-left") # Search
xaringanExtra::use_progress_bar(color = "#0051BA", location = "bottom", 
                                height = "4px")
xaringanExtra::use_clipboard() # Copy Code 
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)
xaringanExtra::use_editable(expires = 1) # Add textboxes to edit during presentation
```

# What is Experimental Design?

- An experimental design is the laying out of a detailed experimental plan in advance of doing the experiment. 

- Why it is important?
    - Well chosen experimental design maximize the amount of information that can be obtained for a given amount of experimental effort. 
    - Without proper design, inferential statistics is something not even possible. 

---
# Why should you care?
- Without proper design it is often impossible to get the information you want from data collected.

- It will save you and time and money.

- Employers know this.. And know they need people who can do it.

---
# Key Information for Experimental Design
1. Define the research question: Hypothesis or study objectives.

2. State variables of interest: Independent and dependent variables, controls, covariates.

3. Construct a sampling design:
    - Define the population of interest.
    - Determine the sampling unit: What is an observation?
    - Appropriate sampling unit size and shape.
    - How many observations are needed?
    - Selected and position of observations: consider control and pairing.

4. Develop your **method**: how, where, when to sample?

- This information will help you to choose the appropriate statistical analysis.
     - We'll cover this throughout the course.

---
# Define the problem: Hypothesis or study objective
- Hypothesis: statement of the expected outcome of an experiment.

- Characteristics of a good Hypothesis:
    - Clear, concise statement of expected results.
    - Based on scientific theory or observation
    - Is testable
    - Identifies the key variables included 
    - Specifies the nature of the relationship or differences expected
    - Define the population of interest.

---
# Hypothesis or study objective

.pull-left[ 
- Clear and Concise: Clearly states the expected outcome.

.medium[
    - **Example:** Higher soil moisture will increase tree growth rates in Vermont forests.]

- Based on Theory or Observation: Grounded in prior knowledge or observations.

.medium[
    - **Example:** Based on the understanding that moisture availability influences tree growth.]

- Testable: Can be evaluated with data collection.
    
.medium[
    - **Example:** Measure tree growth in areas with varying soil moisture levels.]

]
.pull-right[
- Identifies Key Variables: Specifies independent and dependent variables.

.medium[
    - **Example:** Soil moisture (independent) and tree growth rate (dependent).]

- Specifies Relationship: Defines the expected interaction between variables.

.medium[
    - **Example:** Increased soil moisture is expected to enhance tree growth.]

- Defines Population: Specifies the study context or target population.

.medium[
    - **Example:** Mature deciduous trees in Vermont forests.]
]

---
# Hypothesis or **study objective**

- A study objective outlines the purpose of research when there is no specific hypothesis to test statistically. Instead, it focuses on describing, monitoring, or understanding a system. Here are the key characteristics:

1. Clear and Concise Statement of the Goal or Knowledge to Be Gained:  
The objective should clearly articulate what the study aims to achieve or learn.
.medium[
    - **Example:** To assess the effects of varying soil moisture on tree growth in Vermont forests.]

2. Identified Key Variables:  
Clearly identify the main variables that will be measured or observed in the study.
.medium[
    - **Example:** Soil moisture levels and tree growth rates.]

3. Defines the Population of Interest:  
.medium[
    - Specify the specific population or system that the study will focus on.]

---
# Variables of interest

- Independent Variable (Manipulated Variable):
    - This is the variable that is deliberately changed or manipulated in the study. It is the primary focus of the experiment to see how it affects the dependent variable.
    - Example: In a study on tree growth, the independent variable could be soil moisture levels.

- Dependent Variable (Responding Variable):
    - This variable changes in response to the manipulation of the independent variable. It is what you measure to assess the effect of the independent variable.
    - Example: The growth rate of trees, which responds to changes in soil moisture.
    
---
# Variables of interest

- **Control:**
 - Controls are factors that are kept constant throughout the experiment to ensure that the results are due to the manipulation of the independent variable, not other factors. Controls help establish a standard of comparison.
 - Example: Using the same tree species, consistent light exposure, and the same soil type across all test groups.

- **Covariate:**  
- Covariates are factors that could influence the dependent variable but cannot be fully controlled. Including them helps to account for their potential impact in statistical analysis, ensuring more accurate results.
- Example: In a tree growth study, covariates might include temperature, nutrient availability, or pest pressure-factors that can influence growth but are difficult to control completely.

---
# Sample, replicate and pseudo replicate 
- **Sample:**
- A statistical sample is a subset of items which is used to represent the population from which they are drawn.

- It has two connotations (Sokal & Rohlf 2009):
1. **Individual observations**, which are taken in small units.
These smaller sampling units are often, but not necessarily, individuals in the biological sense.

If we take the weight of 100 fish, then the weight of each fish is an individual observation.

---
# Sample, replicate and pseudo replicate 
- **Sample:**
- A statistical sample is a subset of items which is used to represent the population from which they are drawn.

- It has two connotations (Sokal & Rohlf 2009):
2. **A set of individuals** that are examined to estimate the average value.
Each set of individuals are considered as basic sampling units.

- If we wish to measure temperature in a study of ant colonies, where each colony is a basic sampling unit, each temperature reading for one colony is an individual observation, and the sample of observations is the temperature for all the colonies considered.


---
# Sample, replicate and pseudo replicate 

- **Replication:** 
- Replication means having observations replicated at a spatial and temporal scale that coincides with the application of the experimental treatments.
- In statistics, replicates mean different samples that capture random variations of the system.
- Replicas are essential because biological systems are inherently variable, and this is particularly true for ecological systems.

---
# Sample, replicate and pseudo replicate

- **Pseudoreplication:**
Hurlbert (1984) to refer to **the use of inferential statistics to test for treatment effects with data from experiments where either treatments are not replicated (though samples may be) or replicates are not statistically independent.**

Heffner et al (1996) distinguish a pseudoreplicate from a true replicate, which they characterize as **the smallest experimental unit to which a treatment is independently applied**.

---
# Sampling design


---
# Sampling design

- Sampling design proposed method of selecting a subset of the population for measurements.

- Observation (or Sampling Unit):
  - Individual Plants or Animals: Measure specific organisms.
  - Plant or Animal Parts: Sample parts (e.g., leaves, scales).
  - Quadrants: Define and sample areas (e.g., 1m^2).
  - Lines (Transects): Sample along a line across the study area.
  - Points: Collect samples at specific points.
  
---
# Understanding Statistical Power
- **Statistical Power** is the probability that a statistical test will correctly reject a null hypothesis (which states that there is no relationship between the variables of interest) when it is false, meaning that a true effect or relationship exists. 
It reflects the ability of the test to detect a significant result if there is one.

- Power Value: A power of 0.9 means there is a 90% chance of detecting a significant difference if it truly exists, and a 10% chance of missing it (Type II error).
For most studies, a power greater than 0.8 is considered adequate, indicating an 80% chance of detecting a significant result if there is a true effect.

- Statistical power is intricately linked to the number of observations.

---
# Understanding Statistical Power

---
# Sampling design

```{r, echo=FALSE}
mermaid("
graph TB
    A[Sampling techniques]
    A --> B[Non-Probability sampling<br>Chosen based on subjective judgment]
    A --> C[Probability sampling<br>Random selection<br>Equal opportunity]
",
height = '100%', width = '100%')
```

