---
title: "Inferential Statistics"
author: "Pablo E. Gutierrez-Fonseca"
date: 'Fall 2024'
tables: true
graphics: yes
output:   
  beamer_presentation:
    theme: "Boadilla"
    keep_tex: true
header-includes: 
  - \usepackage{graphicx}
  - \logo{\ifnum\thepage>1\hfill\includegraphics[width=1cm]{logo}\fi}
  - \titlegraphic{\includegraphics[width=3cm]{logo}}
  - \newcommand{\theHtable}{\thetable}
---


## Expanding on Hypothesis Testing 

- 1-tailed
  - Hypothesis includes an **expected direction**.

:::::::::::::: {.columns}
::: {.column width="50%"}

\vspace{1cm}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=2, fig.height=2}
library(ggplot2)
# Set the significance level
p <- 0.05

# Create a sequence for the lower tail (left tail of the normal distribution)
tail_low <- seq(-4, qnorm(p), 0.01)
df_tl <- data.frame(x = c(tail_low, qnorm(p)), y = c(dnorm(tail_low), 0))

# Plot the normal distribution
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm) +  # Plot the normal distribution
  geom_polygon(data = df_tl, aes(x, y), fill = "#99CCFF") +  # Fill the lower tail (P(X < x))
  geom_vline(xintercept = qnorm(p), lty = "dashed", lwd = 0.3) +  # Vertical line at the critical value for left tail

  annotate(geom = "text", x = -2.7, y = 0.3, label = 'P<x') +  # Annotate the critical value
  
  theme_classic() +  # Clean theme
  labs(x = "", y = "")  # Remove axis labels

```

\centering

\tiny
- Decrease
- Cooler
- Smaller
- Lower
\

\

:::
::: {.column width="50%"}

\vspace{1cm}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=2, fig.height=2}
# Set the significance level
p <- 0.05

# Create a sequence for the upper tail (one-tailed test)
tail_high <- seq(qnorm(1-p), 4, 0.01)
df_th <- data.frame(x = c(qnorm(1-p), tail_high), y = c(0, dnorm(tail_high)))

# Plot the normal distribution
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm) +  # Plot the normal distribution
  geom_polygon(data = df_th, aes(x, y), fill = "#99CCFF") +  # Fill the upper tail
  geom_vline(xintercept = qnorm(1-p), lty = "dashed", lwd = 0.3) +  # Vertical line at the critical value
  annotate(geom = "text", x = 2.7, y = 0.3, label = 'P>x') +  # Annotate the critical value
  
  theme_classic() +  # Clean theme
  labs(x = "", y = "")  # Remove axis labels

```

\centering

\tiny
- Increase
- Warmer
- Higher
- Expand
\

\

:::
::::::::::::::


-------------------------------------------------------------------------------------------


## Expanding on Hypothesis Testing 

- 1-tailed - hypothesis includes an **expected direction**.

:::::::::::::: {.columns}
::: {.column width="50%"}

\vspace{1cm}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=2, fig.height=2}
library(ggplot2)
# Set the significance level
p <- 0.05

# Create a sequence for the lower tail (left tail of the normal distribution)
tail_low <- seq(-4, qnorm(p), 0.01)
df_tl <- data.frame(x = c(tail_low, qnorm(p)), y = c(dnorm(tail_low), 0))

# Plot the normal distribution
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm) +  # Plot the normal distribution
  geom_polygon(data = df_tl, aes(x, y), fill = "#99CCFF") +  # Fill the lower tail (P(X < x))
  geom_vline(xintercept = qnorm(p), lty = "dashed", lwd = 0.3) +  
  annotate(geom = "text", x = -2.9, y = 0.15, label = sprintf("%.2f", qnorm(p)), size=3) +
  annotate(geom = "text", x = -3.3, y = 0.05, label = "0.05", size=3) +  
  theme_classic() +  # Clean theme
  labs(x = "", y = "")  # Remove axis labels

```


:::
::: {.column width="50%"}

\vspace{1cm}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=2, fig.height=2}
# Set the significance level
p <- 0.05

# Create a sequence for the upper tail (one-tailed test)
tail_high <- seq(qnorm(1-p), 4, 0.01)
df_th <- data.frame(x = c(qnorm(1-p), tail_high), y = c(0, dnorm(tail_high)))

# Plot the normal distribution
ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm) +  # Plot the normal distribution
  geom_polygon(data = df_th, aes(x, y), fill = "#99CCFF") +  # Fill the upper tail
  geom_vline(xintercept = qnorm(1-p), lty = "dashed", lwd = 0.3) +  # Vertical line at the critical value
  annotate(geom = "text", x = 2.7, y = 0.15, label = sprintf("%.2f", qnorm(1-p)), size=3) +  
  annotate(geom = "text", x = 3, y = 0.05, label = "0.05", size=3) +  
  
  
  theme_classic() +  # Clean theme
  labs(x = "", y = "")  # Remove axis labels

```


:::
::::::::::::::

- If your obtained test statistic falls beyond the critical value (lightblue) for your given Alpha threshold = Significant result, **reject the null**.


-------------------------------------------------------------------------------------------

## Expanding on Hypothesis Testing 

- 2-tailed tests: 
  - Have **no expected directionality** hypothesized. 
  - Splits the 5% of the area under the curve that would be considered significant between both tails of the normal distribution curve.
  - Are therefore less powerful tests (more likely to find a significant result).

\centering
```{r, echo=FALSE, message=FALSE, fig.cap="", fig.width=2.5, fig.height=2}
p <- 0.025
tail_low <- seq(-4, qnorm(p), 0.01)
df_tl <- data.frame(x=c(tail_low,qnorm(p)), y =c(dnorm(tail_low),0))
tail_high <- seq(qnorm(1-p), 4, 0.01)
df_th <- data.frame(x=c(qnorm(1-p),tail_high), y=c(0,dnorm(tail_high)))

ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm) +
  geom_polygon(data = df_tl, aes(x,y), fill = "#99CCFF") +
  geom_polygon(data = df_th, aes(x,y), fill = "#99CCFF") +
  geom_vline(xintercept = qnorm(p), lty = "dashed", lwd = 0.3) +
  geom_vline(xintercept = qnorm(1-p), lty = "dashed", lwd = 0.3) +
  annotate(geom = "text", -2.9, 0.15, label = "-1.96", size=3) +
  annotate(geom = "text", 2.7, 0.15, label = "1.96", size=3) +
  
  annotate(geom = "text", -3.3, 0.05, label = "0.025", size=3) +
  annotate(geom = "text",  3, 0.05, label = "0.025", size=3) +
  
  theme_classic() +
  labs(x="", y="")
```
\


-------------------------------------------------------------------------------------------

## Significant or Not?

:::::::::::::: {.columns}
::: {.column width="50%"}

- **Not significant**:
\vspace{1cm}

- Accept the null hypothesis.  

<br>

- There is **no** difference between the sample and population mean.

<br>

- Obtained test statistic < critical value threshold.

<br>

- p-value > alpha threshold (usually 0.05).


:::
::: {.column width="50%"}

**Significant**: 
\vspace{1cm}

- Reject the null hypothesis, 

<br>

- There **is** a difference between the sample and population mean.

<br>

- Obtained test statistic > critical value threshold.

<br>

- p-value < alpha threshold (usually 0.05).


:::
::::::::::::::

-------------------------------------------------------------------------------------------  

## Basic steps for an **Inferential Test**


:::::::::::::: {.columns}
::: {.column width="70%"}

- A statement of null hypothesis. 

<br>

- Choose the appropriate test.

<br>

- Set the level of Type I error risk (alpha)

<br>

- Analyze data distribution 

<br>

- Compute the test statistic (obtained) value

- Assess significance:
  - Determine the critical value needed to  reject the null hypothesis and compare it to your   - calculated test statistic
  - Determine the p-value associated with your calculated test statistic   

<br>

- Summarize


:::
::: {.column width="30%"}

contents...

:::
::::::::::::::

------------------------------------------------------------------------------------------- 

## Basic steps for an **Inferential Test**

- We can select an appropriate test simply by answering some questions.

- What type of data do we have?
  - If we have **frequency data**, we select the Chi-square family.
  - **Continuous and categorical variables** will  have a variety of different tests depending on the research question:
  
- What type of research question are we considering?
  - If **comparing one sample mean** to a **population** = \textcolor{red}{z-test}.
  - If focus is on **differences** we go to the family of tests concerned with comparing groups or treatments (i.e. \textcolor{red}{t-tests, ANOVA}).
  - If the focus is on **relationships** between variables, we go to the \textcolor{red}{correlation tests}.
  - If we want to **predict outcomes** we go to the \textcolor{red}{regression family}.

:::::::::::::: {.columns}
::: {.column width="40%"}

\centering

\( y = \beta_0 + \beta_1 x + \epsilon \)

:::
::: {.column width="60%"}

where:  
- \( y \) is the dependent variable,  
- \( x \) is the independent variable,  
- \( \beta_0 \) is the intercept,  
- \( \beta_1 \) is the slope,  
- \( \epsilon \) is the error term.  

:::
::::::::::::::

------------------------------------------------------------------------------------------- 

## Basic steps for an **Inferential Test**

- We can select an appropriate test simply by answering some questions.

  - How many variables, how many groups are we including? (multivariate)

  - Are observations independent from each other or purposely paired?

  - Is my data not normally distributed? (non-parametric tests)

-------------------------------------------------------------------------------------------

## Statistical shorthand **z-test** example

\centering
\( z_{100} = 1.4, p = 0.16 \)
\

-------------------------------------------------------------------------------------------

## Statistical shorthand **z-test** example

\centering
\( z_{100} = 1.4, \, p = 0.16 \)
\[
\text{\huge $\downarrow$} \quad
\]
\centering
\[
\underbrace{z_{\underbrace{100}_{\text{Sample size}}}}_{\text{Test statistic}} = \underbrace{1.4}_{\text{z-value}}, \quad \underbrace{p = 0.16}_{\text{p-value}}
\]

- Every statistical test has a letter designating the type of test.
- Because power is so closely tied to sample size, we report sample size.
- For clarity (and to confirm the correct-tailed test is used to assess the p-value), include the calculated test statistic value.
- Always report the p-value.
- Some tests will also include a secondary metric to assess how meaningful results are (if significant).

-------------------------------------------------------------------------------------------

## How to summarize your analysis: Key Components of a Statistical Summary

\textcolor{teal}{Before installing a new air quality monitoring instrument, we tested to see if a sample of measurements taken at the testing lab differed significantly different from the long-term mean for the larger network population.}  

<br>

\textcolor{orange}{Using a 2-tailed, one-sample z-test for on our normally distributed samples (W = 0.78).}  

<br>

\textcolor{blue}{While the mean of the new instrument sample was slightly higher (sample mean = 117 vs. population mean = 108), we found no significant difference (z(100) = 1.4, p = 0.16).}    

<br>

\textcolor{darkgray}{Based on these results we approve the installation of this instrument at the new site. However, we would suggest statistical comparisons of the current and new unit side-by-side, prior to decommissioning the current instrument.}

-------------------------------------------------------------------------------------------

## How to summarize your analysis: Key Components of a Statistical Summary

\textcolor{teal}{Introduction/Background:\\
Before installing a new air quality monitoring instrument, we tested to see if a sample of measurements taken at the testing lab differed significantly different from the long-term mean for the larger network population.} 

<br>

\textcolor{orange}{Methods:\\
Using a 2-tailed, one-sample z-test for on our normally distributed samples (W = 0.78).}  

<br>

\textcolor{blue}{Results:\\
While the mean of the new instrument sample was slightly higher (sample mean = 117 vs. population mean = 108), we found no significant difference \( z_{100} = 1.4, \, p = 0.16 \).} 

<br>

\textcolor{darkgray}{Implications:\\
Based on these results we approve the installation of this instrument at the new site. However, we would suggest statistical comparisons of the current and new unit side-by-side, prior to decommissioning the current instrument.}


## 
\vspace{2cm}
\centering
\textcolor{blue}{Our First Inferential Tests!\\
One sample z-test}
\
\
\

-------------------------------------------------------------------------------------------

## When would you run a one-sample z-test?

- Use the One Sample z-test when:
  - You want to test for a difference between one sample mean and a larger population mean.
  - There is only one group (sample) being tested against the larger population.
  - You know (or can estimate) the mean and standard deviation of the population.
  - Data is normally distributed.

\[
z = \frac{\bar{X} - \mu}{\frac{\sigma}{\sqrt{n}}}
\]

where:  
- \( \bar{X} \): Sample mean.  
- \( \mu \): Population mean.  
- \( \sigma \): Population standard deviation.
- \( n \): Sample size.


-------------------------------------------------------------------------------------------

## When would you run a one-sample z-test?

- Use the One Sample z-test when:
  - You want to test for a difference between one sample mean and a larger population mean.
  - There is only one group (sample) being tested against the larger population.
  - You know (or can estimate) the mean and standard deviation of the population.
    - \textcolor{orange}{Expert Information.}
    - \textcolor{orange}{Scientific Literature.}
    - \textcolor{orange}{Data archives.}
    - \textcolor{orange}{Meaningful (hypothesized value).}
  - Data is normally distributed.

-----------------------------------------------------------------------------


## Flowcharts: Keeping it simple with questions

-----------------------------------------------------------------------------

## One sample z-test

Assume you have a sample of **20 observations**. The mean of this **sample is 150**, and you want to determine whether there is a **difference between the sample mean and the larger population mean**. Since you are not specifying a direction for this difference, you will conduct a two-tailed test.

The **population mean is 164**, and the **population standard deviation is 33**.

Next, calculate the obtained value for this **one-sample z-test.**

:::::::::::::: {.columns}
::: {.column width="50%"}
\vspace{1cm}
\[
z = \frac{\bar{X} - \mu}{\frac{\sigma}{\sqrt{n}}}
\]

:::
::: {.column width="50%"}
\vspace{1cm}
\[
z = \frac{150 - 164}{\frac{33}{\sqrt{20}}} = -1.897
\]

:::
::::::::::::::



-----------------------------------------------------------------------------

## One sample z-test

Assume you have a sample of **20 observations**. The mean of this **sample is 150**, and you want to determine whether there is a **difference between the sample mean and the larger population mean**. Since you are not specifying a direction for this difference, you will conduct a two-tailed test.

The **population mean is 164**, and the **population standard deviation is 33**.

Next, calculate the obtained value for this **one-sample z-test.**

:::::::::::::: {.columns}
::: {.column width="50%"}

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=2.3, fig.height=1.8}
p <- 0.025
tail_low <- seq(-4, qnorm(p), 0.01)
df_tl <- data.frame(x=c(tail_low,qnorm(p)), y =c(dnorm(tail_low),0))
tail_high <- seq(qnorm(1-p), 4, 0.01)
df_th <- data.frame(x=c(qnorm(1-p),tail_high), y=c(0,dnorm(tail_high)))

ggplot(data.frame(x = c(-4, 4)), aes(x)) +
  stat_function(fun = dnorm) +
  geom_polygon(data = df_tl, aes(x,y), fill = "#99CCFF") +
  geom_polygon(data = df_th, aes(x,y), fill = "#99CCFF") +
  geom_vline(xintercept = qnorm(p), lty = "dashed", lwd = 0.3) +
  geom_vline(xintercept = qnorm(1-p), lty = "dashed", lwd = 0.3) +
  annotate(geom = "text", -2.9, 0.15, label = "Reject\n Null", size=3) +
  annotate(geom = "text", 2.7, 0.15, label = "Reject\n Null", size=3) +
  
  annotate(geom = "text", -3.3, 0.05, label = "0.025", size=3) +
  annotate(geom = "text",  3, 0.05, label = "0.025", size=3) +
  
  theme_classic() +
  labs(x="", y="")
```

:::
::: {.column width="50%"}
\vspace{1cm}
\[
z = \frac{150 - 164}{\frac{33}{\sqrt{20}}} = -1.897
\]

:::
::::::::::::::


-----------------------------------------------------------------------------

## Critical Value Approach

- In the critical value approach, we compare the calculated test statistic to the critical value from the standard normal distribution. For a two-tailed test at the \( \alpha = 0.05 \) significance level, the critical values are approximately ±1.96, which corresponds to the **0.025** in each tail.

\vspace{1cm}

- Decision Rule

- If \( | \text{calculated test statistic} | > \text{critical value} \) = **Significant**
- If \( | \text{calculated test statistic} | \leq \text{critical value} \) = **Not Significant**


-----------------------------------------------------------------------------

## Critical Value Approach

In this case:

\centering
\[
|-1.897| < 1.96
\]
\

- Thus, we fail to reject the null hypothesis, indicating that there is **no significant** difference between the sample mean and the population mean.

-----------------------------------------------------------------------------

## Finding the p-value

Now we need to determine the probability (p-value) associated with the calculated test statistic. You have several options to achieve this:

```{r}
z_value <- -1.897
p_value <- 2 * pnorm(z_value)  # Two-tailed p-value
p_value  # This will give the result
```

- Since this p-value (0.0578) is greater than the alpha threshold of 0.05, we conclude that the result is **not significant**.

-----------------------------------------------------------------------------

## Example

- A rental car company claims the mean time to rent a car on their website is 60 seconds with a standard deviation of 30 seconds. A random sample 36 customers attempted to rent a car on the website. The mean time to rent was 75 seconds. Is this enough evidence to contradict the company's claim?
\vspace{0.5cm}

Step 1: Set up the hypotheses
\vspace{0.5cm}

:::::::::::::: {.columns}
::: {.column width="50%"}

- **Null hypothesis (\( H_0: \))**: The mean rental time is 60 seconds.
  \[
  H_0: \mu = 60
  \]

:::
::: {.column width="50%"}

- **Alternative hypothesis (\( H_1: \))**: The mean rental time is different from 60 seconds.
  \[
  H_1: \mu \neq 60
  \]
  
:::
::::::::::::::


-----------------------------------------------------------------------------


## Example

- A rental car company claims the mean time to rent a car on their website is 60 seconds with a standard deviation of 30 seconds. A random sample 36 customers attempted to rent a car on the website. The mean time to rent was 75 seconds. Is this enough evidence to contradict the company's claim?
\vspace{0.5cm}

Step 1: Test statistic formula
\vspace{0.5cm}

:::::::::::::: {.columns}
::: {.column width="50%"}


\[
z = \frac{\bar{X} - \mu}{\frac{\sigma}{\sqrt{n}}}
\]

:::
::: {.column width="50%"}

\[
z = \frac{75 - 60}{\frac{30}{\sqrt{36}}} = \frac{15}{5} = 3
\]
  
:::
::::::::::::::


-----------------------------------------------------------------------------

## Example

- A rental car company claims the mean time to rent a car on their website is 60 seconds with a standard deviation of 30 seconds. A random sample 36 customers attempted to rent a car on the website. The mean time to rent was 75 seconds. Is this enough evidence to contradict the company's claim?
\vspace{0.5cm}

Step 2: Set up the hypotheses

:::::::::::::: {.columns}
::: {.column width="40%"}

For a **one-tailed test** at a significance level of \( \alpha = 0.05 \), the critical z-value is approximately **1.645** (since we are only looking at one tail).


:::
::: {.column width="60%"}

```{r}
# Given z-value
z_value <- 3
# Calculate the one-tailed p-value
p_value <- 1 - pnorm(z_value)
# Print the p-value
p_value
```
  
:::
::::::::::::::


-----------------------------------------------------------------------------

## Example: Car rental

\textcolor{teal}{We tested a rental car company's claim that reservations could be made online in 60 seconds on average.} \textcolor{orange}{ Using a random sample of 36 online rentals in a one-tailed, one-sample z-test} \textcolor{blue}{ we found that mean reservation time is actually significantly higher than the company claims (\( z_{(36)}=3, \, p=0.001 \)).} \textcolor{darkgray}{  This result indicates that it is highly unlikely customers can rent a car in 60 seconds as claimed.  We suggest that the rental company adjust its claim to match the 75 second mean encountered for our test sample to avoid misleading potential customers.}

