---
title: "Hypothesis Testing, Probability and Distribution xaringan"
subtitle: ""
author: "Pablo E. Guti\u00E9rrez-Fonseca"
institute: ""
date: "2025-02-12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, "css/nhsr.css", "css/nhsr-fonts.css", "css/custom.css"]
    nature:
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      highlightSpans: true 
      countIncrementalSlides: true
      ratio: "16:9"
    includes:
      after_body: [css/insert-logo.html] #this is for logos in all slides
xaringanExtra:
    use_panelset: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(readr)
library(lessR)
library(ggplot2)
library(patchwork)
library(palmerpenguins)
library(car)
library(ggforce) # for geom_circle
library(RVAideMemoire) #shapiro.test
library(DiagrammeR)
knitr::opts_chunk$set(dpi= 300)
xaringanExtra::use_panelset()
xaringanExtra::use_scribble()
xaringanExtra::use_search(show_icon = FALSE, position= "bottom-left") # Search
xaringanExtra::use_progress_bar(color = "#0051BA", location = "bottom", 
                                height = "4px")
xaringanExtra::use_clipboard() # Copy Code 
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)
xaringanExtra::use_editable(expires = 1) # Add textboxes to edit during presentation
```


# Normal distribution: Introduction
- Why did we focus on normality?

  - The **normal distribution** is a key tool for determining the probability of a given value occurring in a population that follows this distribution.

  - It allows us to make inferences about a population by calculating how likely it is for data to fall within certain ranges.

  - Many statistical tests assume data follows a **normal distribution**, which helps in determining **significance** and making reliable conclusions.

---
## Normal distribution: Introduction
- The normal distribution has two parameters: the mean ( $\mu$) and the standard deviation ( $\sigma$ ).

- If a **variable x** follows a normal distribution with a specific mean ( $\mu$ ) and standard deviation ( $\sigma$ ), we write this as:

.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=1.6, fig.height=1.8}
mean <- 0  # Mean 
sd <- 1    # Standard deviation 
# Generate a sequence of x values
x <- seq(-4, 4, length = 100)
# Calculate the density of the normal distribution for each x value
y <- dnorm(x, mean, sd)
# Create a data frame for ggplot
data <- data.frame(x = x, y = y)
# Plot the normal distribution
ggplot(data, aes(x = x, y = y)) +
  geom_line(color = "blue", size = 1) +
  geom_vline(xintercept = mean, linetype = "dashed", color = "red", size = 1) +
  labs(title = "", subtitle = expression(paste("")),x = "", y = "") +
  theme(axis.text.y = element_blank(), # Remove y-axis numbers
      #  axis.ticks.y = element_blank(), # Remove y-axis ticks
        axis.title.y = element_blank()) +# Remove y-axis title
  theme_minimal()
```
] 
.pull-right[
.center[
$x$ $\sim$ N( $\mu$, $\sigma$ )
]]
