---
title: "Hypothesis Testing, Probability and Distribution xaringan"
subtitle: ""
author: "Pablo E. Guti\u00E9rrez-Fonseca"
institute: ""
date: "2025-02-12 (updated: `r Sys.Date()`)"
output:
  xaringan::moon_reader:
    lib_dir: libs
    css: [default, "css/nhsr.css", "css/nhsr-fonts.css", "css/custom.css"]
    nature:
      highlightLanguage: r
      highlightStyle: github
      highlightLines: true
      highlightSpans: true 
      countIncrementalSlides: true
      ratio: "16:9"
    includes:
      after_body: [css/insert-logo.html] #this is for logos in all slides
xaringanExtra:
    use_panelset: true
---

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(readxl)
library(readr)
library(lessR)
library(ggplot2)
library(cowplot)
library(patchwork)
library(palmerpenguins)
library(car)
library(ggforce) # for geom_circle
library(RVAideMemoire) #shapiro.test
library(DiagrammeR)
knitr::opts_chunk$set(dpi= 300)
xaringanExtra::use_panelset()
xaringanExtra::use_scribble()
xaringanExtra::use_search(show_icon = FALSE, position= "bottom-left") # Search
xaringanExtra::use_progress_bar(color = "#0051BA", location = "bottom", 
                                height = "4px")
xaringanExtra::use_clipboard() # Copy Code 
xaringanExtra::use_extra_styles(
  hover_code_line = TRUE,         #<<
  mute_unhighlighted_code = TRUE  #<<
)
xaringanExtra::use_editable(expires = 1) # Add textboxes to edit during presentation
```


# Normal distribution: Introduction
- Why did we focus on normality?

  - The **normal distribution** is a key tool for determining the probability of a given value occurring in a population that follows this distribution.

  - It allows us to make inferences about a population by calculating how likely it is for data to fall within certain ranges.

  - Many statistical tests assume data follows a **normal distribution**, which helps in determining **significance** and making reliable conclusions.

---
## Normal distribution: Introduction
- The normal distribution has two parameters: the mean ( $\mu$) and the standard deviation ( $\sigma$ ).

- If a **variable x** follows a normal distribution with a specific mean ( $\mu$ ) and standard deviation ( $\sigma$ ), we write this as:

.pull-left[
```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=1.6, fig.height=1.8}
mean <- 0  # Mean 
sd <- 1    # Standard deviation 
# Generate a sequence of x values
x <- seq(-4, 4, length = 100)
# Calculate the density of the normal distribution for each x value
y <- dnorm(x, mean, sd)
# Create a data frame for ggplot
data <- data.frame(x = x, y = y)
# Plot the normal distribution
ggplot(data, aes(x = x, y = y)) +
  geom_line(color = "blue", size = 1) +
  geom_vline(xintercept = mean, linetype = "dashed", color = "red", size = 1) +
  labs(title = "", subtitle = expression(paste("")),x = "", y = "") +
  theme(axis.text.y = element_blank(), # Remove y-axis numbers
      #  axis.ticks.y = element_blank(), # Remove y-axis ticks
        axis.title.y = element_blank()) +# Remove y-axis title
  theme_minimal()
```
] 
.pull-right[
.center[
$x$ $\sim$ N( $\mu$, $\sigma$ )
]]

---
# Converting to Standard Normal Distribution

- To simplify comparisons between different normal distributions, we often convert values to the **standard normal distribution**.

- The standard normal distribution has a $\mu$ = $0$ and a $\sigma$ = $1$.

- Any normal distribution can be converted into the standard normal distribution by using the **z-score** formula: $z = \frac{x - \mu}{\sigma}$

```{r, echo=FALSE, message=FALSE, warning=FALSE, fig.width=2, fig.height=1}
mu <- 2
sd <- 1
x_vals <- seq(mu - 4 * sd, mu + 4 * sd, length.out = 100)
dist1 <- dnorm(x_vals, mean = mu, sd = sd)
data1 <- data.frame(x = x_vals, y = dist1)

x_vals2 <- seq(-4 * sd, 4 * sd, length.out = 100)
dist2 <- dnorm(x_vals2, mean = 0, sd = sd)
data2 <- data.frame(x = x_vals2, y = dist2)

# First plot: Normal distribution with mean mu
plot1 <- ggplot(data1, aes(x = x, y = y)) +
  geom_line(color = "blue", size = 1) +             # Plot the line
  geom_ribbon(data = subset(data1, x >= mu & x <= mu + sd), aes(ymin = 0, ymax = y), fill = "lightblue", alpha = 0.3) + # Fill light blue from mu to 1 SD
  labs(x = "", y = "") +
  scale_x_continuous(breaks = c(mu, mu + sd), labels = c(expression(mu), "x")) + # Label x-axis with mu and x at mu + 1 SD
  theme_minimal() +
  theme(axis.title.x = element_text(size = 14),axis.title.y = element_text(size = 14)
  )

# Second plot: Normal distribution with mean 0
plot2 <- ggplot(data2, aes(x = x, y = y)) +
  geom_line(color = "red", size = 1) +              # Plot the line
  geom_ribbon(data = subset(data2, x >= 0 & x <= 0 + sd), aes(ymin = 0, ymax = y), fill = "lightblue", alpha = 0.3) + # Fill light blue from 0 to 1 SD
  labs(x = "", y = "") +
  scale_x_continuous(breaks = c(0, 0 + sd), labels = c("0", "z")) + # Label x-axis with 0 and z at 1 SD
theme_minimal() +
  theme(axis.title.x = element_text(size = 16),
        axis.title.y = element_text(size = 16),
        axis.text = element_text(size = 14))
combined_plot <- plot_grid(plot1, plot2, labels = c("A", "B"),  label_size = 12)

print(combined_plot)

```

