---
title: "Hypothesis Testing, Probability and Distributions"
author: "Pablo E. Gutiérrez-Fonseca"
date: 'Fall 2024'
tables: true
graphics: yes
output:   
  beamer_presentation:
    theme: "Boadilla"
    keep_tex: true
header-includes: 
  - \usepackage{graphicx}
  - \logo{\ifnum\thepage>1\hfill\includegraphics[width=1cm]{logo}\fi}
  - \titlegraphic{\includegraphics[width=3cm]{logo}}
  - \newcommand{\theHtable}{\thetable}
---



## Normality, Probability and Significance
- Why did we focus on normality?
  - The **normal distribution** is a key tool for determining the probability of a given value occurring in a population that follows this distribution.

<br>

- It allows us to make inferences about a population by calculating how likely it is for data to fall within certain ranges.

<br>

- Many statistical tests assume data follows a **normal distribution**, which helps in determining **significance** and making reliable conclusions.


## Hypothesis Testing 
- All inferential tests use a formula that calculates a **test statistic**, quantifying the relationship or difference you are testing.

::: columns
:::: column
\vspace{1cm}
- Independent t-test:
```{r echo=FALSE, results='asis'}
cat("$$t = \\frac{\\bar{X_1} - \\bar{X_2}}{\\sqrt{\\frac{s_1^2}{n_1} + \\frac{s_2^2}{n_2}}}$$")

```

  - Dependent t-test
```{r echo=FALSE, results='asis'}
cat("$$t = \\frac{\\bar{D}}{\\frac{s_D}{\\sqrt{n}}}$$")
```

  - One sample z-test
```{r echo=FALSE, results='asis'}
cat("$$z = \\frac{\\bar{X} - \\mu}{\\frac{\\sigma}{\\sqrt{n}}}$$")
```
  
::::

:::: column
\vspace{1cm}
- F-test (ANOVA)
```{r echo=FALSE, results='asis'} 
cat("$$F = \\frac{MST}{MSE}$$")
```

- Pearson correlation
```{r echo=FALSE, results='asis'} 
cat("$$r = \\frac{\\sum (x_i - \\bar{x})(y_i - \\bar{y})}{\\sqrt{\\sum (x_i - \\bar{x})^2 \\sum (y_i - \\bar{y})^2}}$$")
```
::::
:::
