---
title: "PS_ Descriptive Statistics for TAs"
author: "Pablo E. Gutiérrez-Fonseca"
date: "2024-02-08"
output: pdf_document
---




# **R practice.<br />**
\vspace{1em}

**Install packages.<br />**  
```{r}
library(tidyverse)
library(lessR)
library(DT)
library(e1071)
```
\vspace{1em}

**Load the water pollution data into R.<br />** 
```{r, echo=FALSE}
df_strenth <- read.csv('Boardstrength.csv')
summary(df_strenth)
```
\newpage

1. the **MEAN Strength**.  
```{r}
round(mean(df_strenth$Strength),2)
```
\vspace{1em}

1.1. the **MEDIAN Strength**.  
```{r}
round(median(df_strenth$Strength),2)
```
\vspace{1em}


2. Now calculate the **STANDARD DEVIATION** for the Strength.  
```{r}
round(sd(df_strenth$Strength),2)
```
\vspace{1em}

2.1
```{r}
round(var(df_strenth$Strength),2)
```


3. What is the **inter-quartile range** for Strength column?
```{r}
# Calculate the IQR for the specified column
round(iqr_value <- IQR(df_strenth$Strength),2)

# Print the IQR
print(iqr_value)
```
\vspace{1em}

4. Using the Interquartile range technique, how many **OUTLIER** years are there in your
**Strenght** column?   
```{r}
# Calculate the quartiles and IQR
q1 <- quantile(df_strenth$Strength, 0.25, na.rm = TRUE)
q3 <- quantile(df_strenth$Strength, 0.75, na.rm = TRUE)
iqr <- q3 - q1

# Define the lower and upper bounds for outliers
lower_bound <- q1 - 1.5 * iqr
upper_bound <- q3 + 1.5 * iqr

# Identify outlier years
outlier_years <- df_strenth$Strength < lower_bound | df_strenth$Strength > upper_bound

# Count the number of outlier years
num_outliers <- sum(outlier_years)

# Print the number of outlier years
print(num_outliers)
```
\vspace{1em}

5. Now, for the **Density** column, calculate the **skewness, standard error of skewness (SES), kurtosis, standard error of kurtosis (SEK)**, and perform the **normality test with Shapiro-Wilk test**. Based on all of these parameters, determine if the Density is normally distributed.   
```{r}
# Calculate Pearson's skew
skew <- skew(df_strenth$Density)
# Print the skew rounded to 2 decimal places
print(round(skew, 2))
```
\vspace{1em}

5.1. Calculate the **standard error of skew (ses)** for this data so we can determine the
significance of our skew. 
```{r}
# Calculate the skewness and standard error of skew
ses <- round(sqrt(6/length(df_strenth$Density)),2)
ses
```
\vspace{1em}

5.2. Based on the ses technique, is this **skew significant**?  
**No**
\vspace{1em}

5.3. Now determine the **kurtosis for the Yearly Mean Snow Depth data**. 
```{r}
kurtosis <- round(kurtosis(df_strenth$Density, type = 1) - 3, 2)
kurtosis

# pivot(df_strenth, c(IQR, skew, kurtosis, mean, sd, var), Density)


```
\vspace{1em}

5.4. Calculate the **standard error of kurtosis (sek)** for this data so we can determine the significance of our skew. 
```{r}
# Calculate standard error of kurtosis (SES)
ses_kurt <- round(sqrt(24/length(df_strenth$Density)),2)
ses_kurt
```
\vspace{1em}

5.6. Based on the sek technique, is this **kurtosis significant**? 
**No**
\vspace{1em}

5.7. Based on all of these parameters, determine if the Yearly Mean Snow Depth is normally distributed. Pay special attention to the shape (distribution) of the data (plotted on a histogram).
**No**
\vspace{1em}

5.8. Is your data normally distributed? How can you tell? Be sure to report your certainty of this conclusion.
```{r}
shapiro.test(df_strenth$Density)
```
\vspace{1em}
 
 
```{r}
hist(df_strenth$Density, breaks = 10)
```
 
