---
title: "PS_ Probability and Hypothesis Testing for TAs"
author: "Pablo E. Gutiérrez-Fonseca"
date: "2023-08-12"
output: pdf_document
---

# **R practice.<br />**
\vspace{1em}

**Install packages.<br />**  
```{r}
library(ggplot2)
library(dplyr)
```
\vspace{1em}

**Load the water pollution data into R.<br />** 
```{r, echo=FALSE}
df_mammals <- read.csv('C:/Users/Guti/OneDrive - University of Vermont/Curriculum/07_ Cursos/NR-2400/Problem set_ M5 Probability and Hypothesis Testing/Sleeping Mammals.csv')
summary(df_mammals)
```
\newpage

1. Based on the summary results in JMP, mammals spend most of their time in which type of
sleep phase (dreaming or non-dreaming)?  
```{r}
mean(df_mammals$NonDreaming, na.rm = TRUE)
mean(df_mammals$Dreaming, na.rm = TRUE)
```
\vspace{1em}

2. Which type of sleep phase has the highest variability across the species included here
(dreaming or non-dreaming)?  
Non-dreaming  
```{r}
sd(df_mammals$NonDreaming, na.rm = TRUE)
sd(df_mammals$Dreaming, na.rm = TRUE)
```
\vspace{1em}


3. Enter the p-value for the goodness of fit test for the **Dreaming variable** (NOTE….just enter the number (no letters, symbols, equal signs etc)…also be careful of decimal places.
```{r}
shapiro.test(df_mammals$Dreaming)
```
\vspace{1em}

4. Based on this goodness of fit test, is the **dreaming variable** normally distributed?
**No**
\vspace{1em}

5. Enter the new p-value for the goodness of fit test for the **Dreaming variable** when the outliers are excluded (NOTE….just enter the number (no letters, symbols, equal signs
etc)…also be careful of decimal places.
```{r}
quartiles <- quantile(na.omit(df_mammals$Dreaming), probs = c(0.25, 0.75))
IQR <- IQR(na.omit(df_mammals$Dreaming))
Lower <- quartiles[1] - 1.5*IQR
Upper <- quartiles[2] + 1.5*IQR 

new_Dreaming <- subset(df_mammals$Dreaming, df_mammals$Dreaming > Lower & df_mammals$Dreaming < Upper)

shapiro.test(new_Dreaming)
```
\vspace{1em}

6. Based on this goodness of fit test, is the dreaming variable normally distributed once
outliers are removed?
**Yes.**  
\vspace{1em}

7. Enter the new p-values for the goodness of fit test for the **log transformed Dreaming variable**.
```{r}
log_Dreaming <- log(df_mammals$Dreaming[df_mammals$Dreaming > 0])
shapiro.test(log_Dreaming)
```
\vspace{1em}

8. Is this **log transformed Dreaming variable** normally distributed?  
**Yes**  
\vspace{1em}

9. Based on these results, would you transform (outlier transform, log transform or no
transform) your **dreaming variable**? Assume that outliers removed are incorrect data
values and that all values fit the range necessary for a log transform) 
**Outlier transform**
\vspace{1em}

10. Now go back and repeat these procedures for the **Non-dreaming variable**. However,
instead of entering answers for every one of these steps, just summarize what you would do
with this variable: Based on these results, would you transform (outlier transform, log
transform or no transform) your non-dreaming variable? Assume that outliers removed
are incorrect data values)
```{r}
shapiro.test(df_mammals$NonDreaming)
```
**no transform**
\vspace{1em}

11. Is the distribution of the Non dreaming variable normally distributed for the small mammalian size group?
**Yes**
\vspace{1em}

**From the df_mammals make two groups (large and small mammals)**
```{r}
df_mammals_large <- df_mammals %>% 
  filter(Mammalian.Size.group == 'large')

mean_NonDreaming_large <- mean(df_mammals_large$NonDreaming, na.rm = TRUE)
sd_NonDreaming_large <- sd(df_mammals_large$NonDreaming, na.rm = TRUE)
```
\vspace{1em}

```{r}
df_mammals_small <- df_mammals %>% 
  filter(Mammalian.Size.group == 'small')

# Calculate mean and standard deviation of the NonDreaming column
mean_NonDreaming_small <- mean(df_mammals_small$NonDreaming, na.rm = TRUE)
sd_NonDreaming_small <- sd(df_mammals_small$NonDreaming, na.rm = TRUE)
```
\vspace{1em}

12. Is the distribution of the Non dreaming variable normally distributed for the large mammalian size group?
Normality test Large Mammals
```{r}
shapiro.test(df_mammals_large$NonDreaming)
```
\vspace{1em}

13. Calculate the z score for the **Baboon** based on its non dreaming measured value for the small class (i.e. use the mean and std from the small class to calculate this z score)
```{r}
mean_NonDreaming_Baboon <- df_mammals$NonDreaming[df_mammals$Species == "Baboon"]

# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Baboon - mean_NonDreaming_small) / sd_NonDreaming_small

# Print the Z-scores
print(z_scores)
```
\vspace{1em}

14. Calculate the z score for the **Patas monkey** based on its non dreaming measured value for the small class (i.e. use the mean and std from the small class to calculate this z score)   
```{r}
mean_NonDreaming_Patas_monkey <- df_mammals$NonDreaming[df_mammals$Species == "Patas monkey"]

# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Patas_monkey - mean_NonDreaming_small) / sd_NonDreaming_small

# Print the Z-scores
print(z_scores)
```
\vspace{1em}

15. Calculate the z score for the **Rhesus monkey** based on its non dreaming measured value for the small class (i.e. use the mean and std from the small class to calculate this z score) 
```{r}
mean_NonDreaming_Rhesus_monkey <- df_mammals$NonDreaming[df_mammals$Species == "Rhesus monkey"]

# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Rhesus_monkey - mean_NonDreaming_small) / sd_NonDreaming_small

# Print the Z-scores
print(z_scores)
```
\vspace{1em}

16. Calculate the z score for the **Roe deer** based on its non dreaming measured value for the small class (i.e. use the mean and std from the small class to calculate this z score) 
```{r}
mean_NonDreaming_Roe_deer <- df_mammals$NonDreaming[df_mammals$Species == "Roe deer"]

# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Roe_deer - mean_NonDreaming_small) / sd_NonDreaming_small

# Print the Z-scores
print(z_scores)
```
\vspace{1em}

17. Based on your knowledge that higher (absolute value) z-scores are more unusual for a given group’s distribution, which class would you assign **Baboon** to (small or large)? (hint compare **Baboon's z scores** for the small and large group……which group does the species z score fit “best” with…..i.e. for which group is that species z score less extreme?
```{r}
# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Baboon - mean_NonDreaming_small) / sd_NonDreaming_small
# Print the Z-scores
print(z_scores)

# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Baboon - mean_NonDreaming_large) / sd_NonDreaming_large
# Print the Z-scores
print(z_scores)
```
**small**
\vspace{1em}


18. Based on your knowledge that higher (absolute value) z-scores are more unusual for a given group’s distribution, which class would you assign **Pata monkey** to (small or large)? (hint compare **Pata monkey's z scores** for the small and large group……which group does the species z score fit “best” with…..i.e. for which group is that species z score less extreme?
```{r}
# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Patas_monkey - mean_NonDreaming_small) / sd_NonDreaming_small
# Print the Z-scores
print(z_scores)

z_scores <- (mean_NonDreaming_Patas_monkey - mean_NonDreaming_large) / sd_NonDreaming_large
# Print the Z-scores
print(z_scores)
```
**small**
\vspace{1em}

19. Based on your knowledge that higher (absolute value) z-scores are more unusual for a given group’s distribution, which class would you assign **Rhesus monkey** to (small or large)? (hint compare **Rhesus monkey's z scores** for the small and large group……which group does the species z score fit “best” with…..i.e. for which group is that species z score less extreme?
```{r}
# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Rhesus_monkey - mean_NonDreaming_small) / sd_NonDreaming_small
# Print the Z-scores
print(z_scores)

z_scores <- (mean_NonDreaming_Rhesus_monkey - mean_NonDreaming_large) / sd_NonDreaming_large
# Print the Z-scores
print(z_scores)
```
**small**
\vspace{1em}

20. Based on your knowledge that higher (absolute value) z-scores are more unusual for a given group’s distribution, which class would you assign **Roe deer** to (small or large)? (hint compare **Roe deer's z scores** for the small and large group……which group does the species z score fit “best” with…..i.e. for which group is that species z score less extreme?
```{r}
# Calculate Z-scores for the NonDreaming column
z_scores <- (mean_NonDreaming_Roe_deer - mean_NonDreaming_small) / sd_NonDreaming_small
# Print the Z-scores
print(z_scores)

z_scores <- (mean_NonDreaming_Roe_deer - mean_NonDreaming_large) / sd_NonDreaming_large
# Print the Z-scores
print(z_scores)
```
**large**
\vspace{1em}

19. How many observations in your data set would be labeled outliers by this IQR definition?
```{r}
quartiles <- quantile(na.omit(df_mammals$NonDreaming), probs = c(0.25, 0.75))
IQR <- IQR(na.omit(df_mammals$NonDreaming ))
Lower <- quartiles[1] - 1.5*IQR
Upper <- quartiles[2] + 1.5*IQR 

new_NonDreaming <- subset(df_mammals$NonDreaming , df_mammals$NonDreaming  > Lower & df_mammals$NonDreaming  < Upper)

# Calculate the number of outliers removed
num_outliers_removed <- length(na.omit(df_mammals$NonDreaming)) - length(new_NonDreaming)

# Print the number of outliers removed
cat("Number of outliers removed:", num_outliers_removed, "\n")
```
\vspace{1em}

20. Armed with your knowledge of what a z score is and how to calculate actual values from given z scores, enter the **upper non-dreaming** value beyond which I may have outliers (values greater than 2 standard deviations from the mean in the positive direction). It may help if you sketch this out so that you can visualize what you are calculating. 
```{r}
# Calculate mean and standard deviation of NonDreaming column
mean_NonDreaming <- mean(df_mammals$NonDreaming, na.rm = TRUE)
sd_NonDreaming <- sd(df_mammals$NonDreaming, na.rm = TRUE)

# Calculate upper bound using z-score (z = 2)
z_score_threshold <- 2
upper_bound <- mean_NonDreaming + (z_score_threshold * sd_NonDreaming)

# Print the upper bound
cat("Upper bound for potential outliers:", upper_bound, "\n")
```
\vspace{1em}

21. Do the same for the **lower non-dreaming value** below which I might consider observations to be outliers based on the fact that they are more than 2 standard deviations under the mean.  
```{r}
# Calculate lower bound using z-score (z = 2)
lower_bound <- mean_NonDreaming - (z_score_threshold * sd_NonDreaming)

# Print the lower bound
cat("Lower bound for potential outliers:", lower_bound, "\n")
```
\vspace{1em}

22. Using this 2-standard deviation threshold, how many potential outliers might be in this **non dreaming** data set?
```{r}
# Calculate the number of potential outliers above the upper bound
num_upper_outliers <- sum(df_mammals$NonDreaming > upper_bound, na.rm = TRUE)

# Calculate the number of potential outliers below the lower bound
num_lower_outliers <- sum(df_mammals$NonDreaming < lower_bound, na.rm = TRUE)

# Calculate the total number of potential outliers
total_num_outliers <- num_upper_outliers + num_lower_outliers

# Print the results
cat("Number of potential outliers above upper bound:", num_upper_outliers, "\n")
cat("Number of potential outliers below lower bound:", num_lower_outliers, "\n")
cat("Total number of potential outliers:", total_num_outliers, "\n")
```
\vspace{1em}

23 and 24. Identify the **upper and lower z-scores** that cumulatively represent < 2% TOTAL chance of occurring. This doesn’t require any calculations, just the determination of two z-scores (upper and lower) from the normal probability table (or excel function).
```{r}
# Find the z-score for the upper 1% cumulative probability (upper bound)
upper_bound_prob <- 0.99
upper_z_score <- qnorm(upper_bound_prob)

# Find the z-score for the lower 1% cumulative probability (lower bound)
lower_bound_prob <- 0.01
lower_z_score <- qnorm(lower_bound_prob)

# Print the results
cat("Upper z-score for < 2% cumulative probability:", upper_z_score, "\n")
cat("Lower z-score for < 2% cumulative probability:", lower_z_score, "\n")
```
\vspace{1em}

25. What is the actual **non-dreaming** value associated with the upper z score identified in the question above.  
```{r}
# Calculate the actual non-dreaming value associated with the upper z-score
actual_upper_value <- mean_NonDreaming + (upper_z_score * sd_NonDreaming)

# Print the result
cat("Actual non-dreaming value associated with the upper z-score:", actual_upper_value, "\n")
```

26. How many **non-dreaming** observations in your data set would be labeled as outliers by this zscore definition? 
```{r}
# Calculate the number of non-dreaming observations labeled as outliers
num_outliers_zscore <- sum(df_mammals$NonDreaming > upper_bound | df_mammals$NonDreaming < lower_bound, na.rm = TRUE)

# Print the result
cat("Number of non-dreaming observations labeled as outliers (z-score definition):", num_outliers_zscore, "\n")
```

