---
title: 'Q1 SLR: beaver dams'
output: pdf_document
date: "2024-12-06"
---

```{r setup, include=FALSE}
library(tidyverse)
library(lessR)
library(readxl)
library(RVAideMemoire)
library(car)
library(DescTools)
library(rstatix)
library(PMCMRplus)
```

# Question 1
In recent decades, beavers have expanded beyond their typical habitats and begun colonizing 
the Arctic tundra. Jones et al. collected data on surface water area and surveyed beaver 
dams to investigate how beaver activity influences hydrology in this region. Beaver dams 
create localized flooding around water bodies, potentially impacting the Arctic's hydrology 
and carbon cycling.

Using data from the Baldwin Peninsula in Alaska, Jones and colleagues analyzed surface 
water area and the number of beaver dams over nearly two decades. In this question, 
evaluate whether the presence of beaver dams is associated with an increase in surface 
water in the tundra.

Perform a linear regression to assess whether there is a significant relationship between 
the number of beaver dams and surface water area. Write a concise paragraph summarizing 
your results and conclusions, ensuring you clearly interpret the regression coefficients 
and significance.


```{r}
data <- read.csv('Q1-SLR-beaver-dams.csv')
head(data)
```


## Set up regression
```{r pressure, echo=TRUE}
mod <- lm(area.ha ~ dams.n, data)
summary(mod)
```


# Get standardized residuals
```{r}
dam.res <- rstandard(mod)
shapiro.test(dam.res)
```

# The Breusch-Pagan test to assess homoscedasticity
```{r}
ncvTest(mod)
```

#Cooks D
```{r}
ggplot(as_tibble(cooks.distance(mod)), aes(value)) + geom_boxplot()
plot(mod,which=4)
```

# Make plot of beaver dams and surface water
```{r}
plot(area.ha ~ dams.n, data, 
     pch = 19, 
     col = "royalblue4",
     ylab = "Surface water area (ha)",
     xlab =  "Number of beaver dams")
#add regression line
#make line width thicker
abline(mod, lwd=2)
```


# References:
  
https://bookdown.org/kroppheather/IntEnvData/introduction-to-linear-regression.html
https://github.com/kroppheather/Intro_Enviro_Data/blob/main/IntroEnvData/data/tutorial_5/beaver_dam.csv


Response Key:&nbsp;

Hypothesis (3 pts): We tested whether the presence of beaver dams is significantly associated with an increase in surface water area in the Arctic tundra.

Test (considering assumptions) (8 pts): We used a linear regression to assess the relationship between the number of beaver dams and surface water area. Assumptions for linear regression, including linearity and homoscedasticity, were considered reasonable based on visual inspection of residuals.

correct values (5 pts): slope = slope = 0.318, p= 0.0027, r2 = 0.610

Description (2 pts): The relationship between beaver dams and surface water area is statistically significant and positive. For each additional beaver dam, surface water area increases by approximately 0.318 hectares on average.

Interpretation/Conclusions (2 pts): The findings suggest that beaver activity, indicated by the number of dams, contributes to an increase in surface water area in the Arctic tundra. This supports the hypothesis that beaver dams influence hydrology in this region, potentially affecting broader ecological processes such as carbon cycling.
