---
title: "M7 Quiz Independent t-test"
author: "Pablo E. Gutiérrez-Fonseca"
date: "`r format(Sys.time(), '%Y-%m-%d %H:%M:%S')`"
output: pdf_document
---

Download **insect_data.xls** data and load it into R for analysis.  

Note that this data contains mean insect density for two different pest species (Group A and B), before and after application of a fertilizer mix. Be sure to view the data frame before conducting any analysis to be sure you understand the data structure.  

You've seen this data in the previous quiz. Now repeat your analyses with this same data to see if there is a significant difference between species A and B in the AFTER data column.  

Summarize the results of this test in one very concise paragraph.

#1. Import libraries and load packages
```{r message=FALSE}
library(tidyverse)
library(lessR)
library(readxl)
library(BSDA)
```


#2. importing our data
```{r}
insect_data <- read_excel("insect_data.xlsx", sheet='indep')
```


```{r}
t.test(After ~ Group, data=data,
       alternative = c("two.sided"),
       mu = 0, 
       paired = FALSE, 
       var.equal = TRUE,
       conf.level = 0.95)
```

```{r}
mean_A <- mean(insect_data$After[insect_data$Group == "A"])
mean_B <- mean(insect_data$After[insect_data$Group == "B"])

sd_after <- sd(insect_data$After)

# Calculate Cohen's d
abs(mean_A - mean_B)/sd_after
```

